% Paper 9: Neural Engine Training for SLO-Aware Agents
%
% =============================================================================
% STATUS: FIRST DRAFT — TBD values pending ANE experiments
% Last updated: 2026-03-03
% =============================================================================
%
% Heterogeneous compute: ANE for rollout inference + MLX/Metal GPU for gradients
% Models: Qwen3.5-0.8B, 2B, 4B (Gated DeltaNet architecture)
% Hardware: M2 Max 64GB (38-core GPU, 16-core ANE, 12-core CPU)
% Tasks: T1–T6 (same 6 structured output tasks as P3/P5/P6)
% 3 SLO tiers: Interactive (2s), Standard (5s), Relaxed (10s)
% Key hypothesis: ANE at ~3W can match MLX throughput at ~30W for rollouts
% =============================================================================
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage[table]{xcolor}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric, calc, matrix}
\usepackage{pgfplots}
\pgfplotsset{compat=1.17}
\usepackage{multirow}

\usepackage[numbers]{natbib}
\usepackage{xurl}
\usepackage{hyperref}
\hypersetup{hidelinks,breaklinks=true}

\newcommand{\pninetyfive}{\ensuremath{\mathrm{p95}}}
\newcommand{\pninetynine}{\ensuremath{\mathrm{p99}}}
\newcommand{\slo}{\ensuremath{\mathrm{SLO}}}
\newcommand{\sslo}{\ensuremath{\mathrm{S@SLO}}}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\need}[1]{\textcolor{red}{\textbf{[#1]}}}

\definecolor{aneorange}{RGB}{255,149,0}
\definecolor{mlxblue}{RGB}{0,122,255}
\definecolor{hybridgreen}{RGB}{52,199,89}
\definecolor{syncred}{RGB}{255,59,48}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  xleftmargin=0pt,
  showstringspaces=false
}

\title{Neural Engine Training for SLO-Aware Agents:\\Heterogeneous Compute on Apple Silicon}

\author{
V. Michael Maloney \\
Neuralift; University of New Hampshire \\
\texttt{mike@neuralift.ai}
}

\date{}

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\cfoot{\thepage}

\begin{document}
\maketitle
\begin{center}
\textit{Preprint. Work in progress.}
\end{center}

% =============================================================================
% ABSTRACT
% =============================================================================
\begin{abstract}
GRPO training for SLO-aware agents spends the majority of wall-clock time on rollout generation---the inference phase where the model samples completions for policy gradient estimation. Papers~II and~IV characterized this on CUDA; Paper~VI replicated it on MLX. In both cases, a single compute unit handles both inference and gradient computation, leaving the other accelerators on the same die idle.

We demonstrate a heterogeneous compute architecture that offloads rollout inference to Apple's Neural Engine (ANE) while gradient computation remains on the Metal GPU, both on the same Apple Silicon chip. Using Qwen3.5 models (0.8B, 2B, 4B) and our previously established AgentSLO-Bench suite (6 structured output tasks, 3 SLO tiers), we characterize ANE inference performance for structured output generation for the first time.

\begin{center}
\small
\begin{tabular}{lcc ccc}
\toprule
\textbf{Model} & \textbf{Compute} & \textbf{tok/s} & \textbf{Power (W)} & \textbf{J/tok} & \textbf{S@SLO$_{2s}$} \\
\midrule
Qwen3.5-0.8B & \textcolor{mlxblue}{MLX/GPU} & \need{TBD} & ${\sim}$30 & \need{TBD} & \need{TBD} \\
Qwen3.5-0.8B & \textcolor{aneorange}{ANE} & \need{TBD} & ${\sim}$3 & \need{TBD} & \need{TBD} \\
Qwen3.5-2B & \textcolor{mlxblue}{MLX/GPU} & \need{TBD} & ${\sim}$30 & \need{TBD} & \need{TBD} \\
Qwen3.5-2B & \textcolor{aneorange}{ANE} & \need{TBD} & ${\sim}$3 & \need{TBD} & \need{TBD} \\
Qwen3.5-4B & \textcolor{mlxblue}{MLX/GPU} & \need{TBD} & ${\sim}$30 & \need{TBD} & \need{TBD} \\
Qwen3.5-4B & \textcolor{aneorange}{ANE} & \need{TBD} & ${\sim}$3 & \need{TBD} & \need{TBD} \\
\bottomrule
\end{tabular}
\end{center}

Our experiments on M2~Max show that ANE achieves \need{TBD: competitive or reduced} token throughput at approximately 3\,W versus 30\,W for MLX/GPU inference---a \need{TBD}$\times$ power efficiency gain on the inference-dominant training phase. We design and evaluate a hybrid GRPO training loop where ANE generates rollouts and MLX computes policy gradients, with weight synchronization via CoreML re-conversion as the key bottleneck. Results show \need{TBD: whether hybrid training converges, throughput comparison, weight sync overhead}. This work extends the deployment gap analysis from CUDA (Papers~I--IV) through MLX (Papers~V--VI) to the Neural Engine, completing the Apple Silicon compute substrate coverage and establishing power efficiency as a first-class deployment criterion.
\end{abstract}

% =============================================================================
\section{Introduction}
\label{sec:introduction}
% =============================================================================

There are three compute units on every Apple Silicon chip. The CPU runs the operating system. The Metal GPU---38 cores on an M2~Max---handles machine learning. And the Neural Engine sits there, 16 cores drawing roughly 3 watts, doing nothing useful for LLM training.

This paper asks whether it should.

Papers~I through~VI established a consistent finding: the deployment gap between accuracy and production success persists across hardware substrates and model architectures~\citep{maloney2025deploymentgap,maloney2025capacity,maloney2025moe,maloney2025edge}. Paper~I found $\rho = 0.09$ on CUDA. Paper~V found $\rho = 0.067$ on MLX. GRPO training can close this gap---Papers~II and~IV showed how composite SLO-aware rewards reshape model behavior---but training is expensive, and the dominant cost is inference. GRPO's group sampling generates $G$ completions per prompt per step. For $G = 4$ and 200 training steps, that is 800 inference calls before a single gradient is computed.

Consider what this means in practice. A solo developer on an M2~Max MacBook Pro runs Paper~VI's MLX training loop. The Metal GPU generates rollouts at 30\,W, then computes gradients at 30\,W, sequentially. The Neural Engine---16 TOPS of dedicated neural network silicon---sits idle the entire time. This is not a theoretical inefficiency. It is silicon going to waste on every single training step.

Two recent open-source projects change the calculus. Anemll~\citep{anemll2024} demonstrated that full LLM inference can run on the Neural Engine via CoreML model conversion, achieving token generation on hardware designed for neural network matrix operations rather than general-purpose GPU compute. Separately, maderix~\citep{maderix2025} showed that backpropagation itself---not just inference---can execute on ANE, though at 11.2\% utilization and limited to single-layer experiments. Together, these projects establish that the Neural Engine is not limited to vision and audio tasks. It can run language models.

The opportunity is specific and measurable. GRPO training has two distinct computational phases: inference (rollout generation) and optimization (gradient computation). These phases have different requirements. Inference is embarrassingly parallel across group members, memory-bound, and does not require gradient tracking. Gradient computation requires autodiff support, optimizer state, and flexible tensor operations. The Metal GPU excels at the latter. The Neural Engine may excel at the former---at a fraction of the power.

\paragraph{What we contribute.}
Four things, each building on the last:

\begin{enumerate}[leftmargin=12pt]
    \item \textbf{The first ANE inference characterization for structured output tasks.} We evaluate Qwen3.5 models (0.8B, 2B, 4B) on the same 6 tasks and 3 SLO tiers used across Papers~III, V, and~VI, measuring token throughput, latency percentiles, JSON validity, and Success@SLO---all on the Neural Engine via CoreML.

    \item \textbf{Per-component power profiling of Apple Silicon during LLM workloads.} Using \texttt{powermetrics}, we measure CPU, GPU, and ANE power draw separately during inference and training, establishing that ANE operates at approximately 3\,W versus the GPU's approximately 30\,W for equivalent inference tasks.

    \item \textbf{A hybrid GRPO architecture exploiting heterogeneous compute.} We implement a four-phase training loop---ANE rollout, reward scoring, MLX gradient computation, weight synchronization---that uses both accelerators on the same chip for their respective strengths.

    \item \textbf{A weight synchronization bottleneck analysis.} The critical path in hybrid training is propagating LoRA weight updates from MLX back to the CoreML model on ANE. We measure this overhead, characterize its scaling with model size, and identify the break-even point where hybrid training outperforms pure-MLX.
\end{enumerate}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    font=\small,
    box/.style={draw, rounded corners, minimum width=2.8cm, minimum height=0.8cm, align=center},
    ane/.style={box, fill=aneorange!25, draw=aneorange!70},
    mlx/.style={box, fill=mlxblue!20, draw=mlxblue!70},
    sync/.style={box, fill=syncred!20, draw=syncred!70},
    reward/.style={box, fill=gray!15, draw=gray!60},
    arrow/.style={-Latex, thick}
]
% Phase 1: ANE Rollout
\node[ane] (rollout) at (0, 0) {Phase 1\\ANE Rollout};
\node[font=\footnotesize, text=aneorange!80!black, below=0.1cm of rollout] {${\sim}$3\,W};

% Phase 2: Reward
\node[reward] (reward) at (3.5, 0) {Phase 2\\Reward Score};
\node[font=\footnotesize, text=gray!60!black, below=0.1cm of reward] {CPU};

% Phase 3: MLX Gradient
\node[mlx] (grad) at (7, 0) {Phase 3\\MLX Gradient};
\node[font=\footnotesize, text=mlxblue!80!black, below=0.1cm of grad] {${\sim}$30\,W};

% Phase 4: Weight Sync
\node[sync] (sync) at (10.5, 0) {Phase 4\\Weight Sync};
\node[font=\footnotesize, text=syncred!80!black, below=0.1cm of sync] {bottleneck};

% Arrows
\draw[arrow] (rollout) -- (reward);
\draw[arrow] (reward) -- (grad);
\draw[arrow] (grad) -- (sync);
\draw[arrow, dashed] (sync) to[bend right=35] (rollout);

% Title
\node[font=\small\bfseries, above=0.6cm of reward.north east] {Hybrid ANE+MLX GRPO Training Loop};
\end{tikzpicture}
\caption{\textbf{Heterogeneous GRPO on Apple Silicon.} The training loop splits across two accelerators: the Neural Engine generates rollouts at ${\sim}$3\,W (Phase~1), while the Metal GPU computes policy gradients at ${\sim}$30\,W (Phase~3). Weight synchronization---exporting LoRA parameters from MLX and re-converting to CoreML format---is the bottleneck (Phase~4). Reward scoring runs on CPU.}
\label{fig:hybrid-loop}
\end{figure}

The structure of this paper follows the series convention. Section~\ref{sec:background} surveys Neural Engine architecture, the Anemll and maderix projects, and heterogeneous computing on SoCs. Section~\ref{sec:methodology} describes our ANE inference provider, evaluation protocol, hybrid GRPO architecture, and power measurement approach. Section~\ref{sec:setup} details the experimental configuration. Section~\ref{sec:results} presents four research questions: ANE inference performance, power efficiency, hybrid GRPO convergence, and the weight synchronization bottleneck. Section~\ref{sec:discussion} contextualizes these findings within the broader compute substrate progression. Sections~\ref{sec:limitations} and~\ref{sec:conclusion} close with limitations and conclusions.

% =============================================================================
\section{Background and Related Work}
\label{sec:background}
% =============================================================================

% ---------------------------------------------------------------------------
\subsection{Apple Neural Engine Architecture}
\label{sec:bg-ane}
% ---------------------------------------------------------------------------

Apple's Neural Engine (ANE) is a fixed-function neural network accelerator integrated into every Apple Silicon chip since the A11 Bionic (2017). The M2~Max variant includes 16 Neural Engine cores with a combined throughput of 15.8 trillion operations per second (TOPS)~\citep{apple2022m2max}. Unlike the Metal GPU, which executes arbitrary compute shaders, the ANE is designed specifically for neural network inference: matrix multiplications, convolutions, and activation functions on quantized integer or low-precision floating-point tensors.

The programming model is indirect. Developers do not write ANE kernels. Instead, they compile models to CoreML format using Apple's \texttt{coremltools} framework~\citep{apple2023coreml}, and the CoreML runtime decides at inference time which operations execute on ANE versus GPU versus CPU. This opacity has made the ANE effectively invisible to the machine learning research community. Until recently, it was used almost exclusively for on-device vision and audio tasks---face detection, speech recognition, image segmentation---where Apple's own applications dominate.

The ANE's power profile is its distinguishing characteristic. While the M2~Max GPU draws 25--35\,W under sustained compute loads, ANE inference typically consumes 2--5\,W~\citep{apple2022m2max}. This 10$\times$ power gap is not a coincidence---it reflects the architectural trade-off between programmability and efficiency. The GPU can execute any computation; the ANE can execute a restricted set of neural network operations at dramatically lower energy cost.

Three constraints limit ANE's applicability. First, the CoreML compilation pipeline imposes a maximum of 119 unique compiled model functions---a hard limit that restricts model partitioning strategies~\citep{anemll2024}. Second, ANE requires static tensor shapes, which conflicts with the variable-length nature of autoregressive text generation. Third, the lack of public documentation for ANE's instruction set means that all optimization must proceed through CoreML's opaque compilation layer.

% ---------------------------------------------------------------------------
\subsection{Anemll: Full LLM Inference on Neural Engine}
\label{sec:bg-anemll}
% ---------------------------------------------------------------------------

Anemll~\citep{anemll2024} is the first open-source project to achieve complete LLM inference on the Apple Neural Engine. The approach is architecturally specific: it decomposes transformer models into CoreML-compatible components---embedding, FFN chunks, and LM head---each compiled separately and chained at inference time through a stateful KV cache managed in CoreML's state API.

The key technical insight is \emph{chunked conversion}. A transformer model's repeated layers are partitioned into chunks (typically 4--8 layers per chunk), each converted to a separate \texttt{.mlmodelc} file. This works around ANE's compilation limits by keeping each chunk below the operation complexity threshold. Anemll uses look-up table (LUT) quantization---\texttt{lut1} for embeddings, \texttt{lut2} for FFN weights, \texttt{lut3} for the LM head---to reduce model size and improve ANE throughput, since the Neural Engine's integer units are faster than its floating-point paths.

For our purposes, Anemll provides the inference substrate. We use its conversion pipeline to produce CoreML models from Qwen3.5 HuggingFace checkpoints, then load them via \texttt{coremltools.models.CompiledMLModel} with \texttt{ComputeUnit.CPU\_AND\_NE} to target the Neural Engine.

% ---------------------------------------------------------------------------
\subsection{Training on the Neural Engine}
\label{sec:bg-maderix}
% ---------------------------------------------------------------------------

The maderix/ANE project~\citep{maderix2025} represents the first public attempt at \emph{training} on the Apple Neural Engine. The approach reverse-engineers ANE's gradient computation capabilities through CoreML's private APIs, achieving backpropagation on single transformer layers.

The results are instructive in their limitations. Peak ANE utilization reaches only 11.2\%, compared to $>$90\% for well-optimized Metal GPU compute. The implementation is restricted to single-layer training---full model training would require chaining gradients across chunks, which interacts poorly with ANE's stateful computation model. Nevertheless, the project demonstrates that gradient computation is not architecturally impossible on ANE; it is merely impractical at current utilization levels.

This finding directly motivates our heterogeneous approach. If pure ANE training achieves 11.2\% utilization, the practical path forward is not to force the entire training loop onto ANE, but to use ANE for what it does well---inference---and the GPU for what it does well---gradient computation.

% ---------------------------------------------------------------------------
\subsection{GRPO and SLO-Aware Training}
\label{sec:bg-grpo}
% ---------------------------------------------------------------------------

Group Relative Policy Optimization (GRPO)~\citep{grpo} is a reinforcement learning algorithm for language model fine-tuning that replaces the value network in PPO~\citep{schulman2017ppo} with group-relative advantages. For each training prompt, the model generates $G$ completions. Each completion receives a reward; advantages are computed relative to the group mean. The policy gradient update is:

\begin{equation}
\nabla_\theta \mathcal{L} = -\frac{1}{G} \sum_{g=1}^{G} \hat{A}_g \nabla_\theta \log \pi_\theta(y_g \mid x) + \beta \nabla_\theta D_{\mathrm{KL}}[\pi_\theta \| \pi_{\mathrm{ref}}]
\label{eq:grpo}
\end{equation}

where $\hat{A}_g = r_g - \bar{r}$ is the group-relative advantage for completion $g$, $\pi_\theta$ is the policy, $\pi_{\mathrm{ref}}$ is the frozen reference model, and $\beta$ controls KL regularization.

Papers~II and~IV established composite SLO-aware rewards for this training loop: $r = w_{\mathrm{json}} r_{\mathrm{json}} + w_{\mathrm{acc}} r_{\mathrm{acc}} + w_{\mathrm{faith}} r_{\mathrm{faith}} - \lambda \max(0, t - T_{\mathrm{SLO}}) - \mu c$, where $r_{\mathrm{json}}$ is schema validity, $r_{\mathrm{acc}}$ is task accuracy, $r_{\mathrm{faith}}$ is faithfulness, $\lambda$ penalizes latency violations, and $\mu$ penalizes token cost~\citep{maloney2025capacity,maloney2025dynamics}.

The computational bottleneck is rollout generation. Each training step requires $G$ forward passes through the model---inference only, no gradients needed. The gradient computation itself is a single forward-backward pass on the best or all completions. For $G = 4$, the inference-to-gradient compute ratio is approximately 4:1 in forward passes, and inference dominates wall-clock time because it runs autoregressively (token by token) while gradient computation benefits from full-sequence parallelism.

% ---------------------------------------------------------------------------
\subsection{The Deployment Gap Across Substrates}
\label{sec:bg-gap}
% ---------------------------------------------------------------------------

Papers~I through~VIII established the deployment gap as a substrate-independent phenomenon:

\begin{itemize}[leftmargin=12pt]
    \item \textbf{Paper~I}~\citep{maloney2025deploymentgap}: On CUDA/RTX~4090, Spearman $\rho = 0.09$ between accuracy and S@SLO at 2s across 13 models. Twelve of thirteen achieved 0\% S@SLO.
    \item \textbf{Paper~V}~\citep{maloney2025moe}: On MLX/Apple Silicon, $\rho = 0.067$ across 9 models including MoE. Hardware changed; the gap did not.
    \item \textbf{Paper~VI}~\citep{maloney2025edge}: GRPO training transfers from CUDA to MLX---capacity thresholds, forgetting patterns, and curve taxonomy are model-intrinsic, not hardware artifacts.
    \item \textbf{Paper~VIII}~\citep{maloney2025standards}: Proposed \texttt{criteria.yaml} certification standard with power efficiency as an evaluation dimension.
\end{itemize}

This paper extends the substrate coverage to the third accelerator on Apple Silicon: the Neural Engine. If the deployment gap is truly model-intrinsic, it should manifest identically regardless of whether inference runs on GPU cores, Metal shaders, or ANE matrix units. We test this.

% ---------------------------------------------------------------------------
\subsection{Heterogeneous Computing on System-on-Chip}
\label{sec:bg-heterogeneous}
% ---------------------------------------------------------------------------

Heterogeneous computing on SoCs is well-established in mobile and embedded systems~\citep{singh2013heterogeneous}. ARM's big.LITTLE architecture routes compute-intensive tasks to high-performance cores and background tasks to efficiency cores. Apple Silicon extends this to neural network accelerators: the CPU, GPU, and ANE share a unified memory architecture with a single address space, eliminating the PCIe transfer overhead that plagues discrete GPU setups.

For machine learning, heterogeneous execution has been explored primarily in the inference context---model parallelism across CPU and GPU~\citep{aminabadi2022deepspeed}, or offloading specific layers to specialized hardware~\citep{sheng2023flexgen}. Heterogeneous \emph{training}---splitting the training loop across different accelerator types---is largely unexplored, in part because discrete GPU systems make inter-accelerator data movement expensive.

Apple Silicon's unified memory changes this equation. Moving a tensor from GPU to ANE does not require a DMA transfer across a bus. The data already lives in shared memory. The cost of heterogeneous training reduces to model format conversion (MLX weights $\to$ CoreML weights), not data movement. This is the specific property we exploit.


% =============================================================================
\section{Methodology}
\label{sec:methodology}
% =============================================================================

% ---------------------------------------------------------------------------
\subsection{ANE Inference Provider}
\label{sec:method-ane-provider}
% ---------------------------------------------------------------------------

We implement an ANE inference provider (\texttt{ane\_local.py}) that presents the same API as our MLX inference provider from Paper~V. The provider loads Anemll-converted CoreML models via \texttt{coremltools.models.CompiledMLModel} with \texttt{ComputeUnit.CPU\_AND\_NE}, which targets the Neural Engine for supported operations and falls back to CPU for unsupported ones.

The provider supports two model layouts:

\paragraph{Chunked layout.} The model is decomposed into separate CoreML files: \texttt{embed.mlmodelc} (token embedding), $N$ \texttt{ffn\_chunk\_*.mlmodelc} files (transformer layer groups), and \texttt{lmhead.mlmodelc} (output projection). Each chunk maintains independent KV cache state via CoreML's stateful prediction API. Token generation proceeds sequentially: embed $\to$ chunk$_0$ $\to$ chunk$_1$ $\to \cdots \to$ chunk$_{N-1}$ $\to$ lmhead $\to$ sample. This layout is required for models that exceed ANE's per-function compilation limits.

\paragraph{Monolithic layout.} A single \texttt{.mlmodelc} file contains the entire model. KV cache state is managed internally. This layout is simpler but limited to smaller models that fit within ANE's compilation constraints.

Both layouts use identical autoregressive generation: prefill processes all input tokens in a single forward pass with a causal attention mask, then decode generates one token at a time with incremental position IDs and mask updates. We measure time-to-first-token (TTFT) separately from total generation latency to characterize the prefill/decode split on ANE.

Prompt construction follows the same schema-injection approach as Papers~V and~VI: the JSON schema is converted to an example object and appended to the prompt with explicit ``respond with only JSON'' instructions. The provider parses the model's raw text output using a three-stage extraction pipeline (direct parse, markdown code block extraction, regex-based JSON object detection) and backfills required schema fields with type-appropriate defaults.

% ---------------------------------------------------------------------------
\subsection{Evaluation Protocol}
\label{sec:method-eval}
% ---------------------------------------------------------------------------

We evaluate on the same 6 tasks and 3 SLO tiers established in Paper~III (AgentSLO-Bench)~\citep{maloney2025agentslobench} and replicated in Papers~V and~VI:

\begin{center}
\small
\begin{tabular}{llcl}
\toprule
\textbf{ID} & \textbf{Task} & \textbf{Prompts} & \textbf{Cognitive Demand} \\
\midrule
T1 & CLINC intent classification & 500 & Low (lookup) \\
T2 & HotpotQA grounded QA & 1000 & Medium (reasoning) \\
T3 & Tool calling & 500 & Medium (structured) \\
T4 & BFCL function routing & 500 & Medium (selection) \\
T5 & SWE code patching & 300 & High (generation) \\
T6 & GSM8K math reasoning & 200 & High (chain-of-thought) \\
\bottomrule
\multicolumn{4}{l}{\footnotesize 3{,}000 prompts per model. Same splits as Papers~III, V, VI.}
\end{tabular}
\end{center}

SLO tiers remain unchanged:

\begin{itemize}[leftmargin=12pt]
    \item \textbf{Interactive} ($T_{\mathrm{SLO}} = 2$\,s): Chat-like responsiveness. The tier where the deployment gap is starkest.
    \item \textbf{Standard} ($T_{\mathrm{SLO}} = 5$\,s): API-level service. The practical minimum for production agents.
    \item \textbf{Relaxed} ($T_{\mathrm{SLO}} = 10$\,s): Batch or background processing. The generous upper bound.
\end{itemize}

Success@SLO is computed identically to all prior papers:
\begin{equation}
\sslo(T) = \frac{1}{N} \sum_{i=1}^{N} \mathbb{1}[\text{correct}_i \wedge t_i \leq T]
\label{eq:sslo}
\end{equation}
where $\text{correct}_i$ indicates that request $i$ produced valid JSON matching the schema \emph{and} the correct task answer, and $t_i$ is the end-to-end latency including any parsing overhead.

The critical addition for this paper is \textbf{power measurement} as a first-class metric. For each evaluation run, we record per-component power draw (CPU, GPU, ANE) and compute energy per token:
\begin{equation}
E_{\mathrm{tok}} = \frac{\bar{P} \cdot t_{\mathrm{total}}}{N_{\mathrm{tokens}}}
\label{eq:energy-per-token}
\end{equation}
where $\bar{P}$ is mean power draw during generation and $N_{\mathrm{tokens}}$ is total tokens generated.

% ---------------------------------------------------------------------------
\subsection{Hybrid GRPO Architecture}
\label{sec:method-hybrid}
% ---------------------------------------------------------------------------

The hybrid trainer (\texttt{ane\_grpo\_adapter.py}) implements a four-phase training loop that splits computation across ANE and Metal GPU:

\paragraph{Phase 1: ANE Rollout (${\sim}$3\,W).} For each training step, a task is sampled from the dataset. The ANE inference provider generates $G = 4$ completions per prompt, each autoregressively. Timing is captured as \texttt{ane\_rollout\_ms}. The Neural Engine handles all forward computation; no gradient graph is constructed.

\paragraph{Phase 2: Reward Scoring (CPU).} Each completion is scored using the composite reward function from Paper~II: schema validity (binary), task accuracy (0--1), latency penalty ($\lambda \max(0, t - T_{\mathrm{SLO}})$), and cost penalty ($\mu c$). Group-relative advantages $\hat{A}_g = r_g - \bar{r}$ are computed for GRPO.

\paragraph{Phase 3: MLX Gradient Computation (${\sim}$30\,W).} The prompt and all $G$ completions are tokenized and fed through the MLX model (with LoRA adapters applied to the last $L$ transformer layers). Policy gradient loss with KL penalty is computed via \texttt{nn.value\_and\_grad}. The Adam optimizer updates LoRA parameters. Timing is captured as \texttt{mlx\_gradient\_ms}.

\paragraph{Phase 4: Weight Synchronization (bottleneck).} Every $K$ steps (\texttt{batch\_update\_interval}), LoRA weights are exported from the MLX model. In principle, these weights must be merged into the base model and re-converted to CoreML format for ANE. In practice, full CoreML re-conversion is the dominant cost---we measure export time as \texttt{weight\_sync\_ms} and analyze its amortization across different values of $K$.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    font=\small,
    every node/.style={align=center},
]
% Timeline
\draw[-Latex, thick] (0, 0) -- (12, 0) node[right] {time};

% Phase blocks
\fill[aneorange!30] (0.2, 0.3) rectangle (3.5, 1.2);
\node[font=\footnotesize] at (1.85, 0.75) {ANE Rollout\\($G{=}4$ samples)};

\fill[gray!20] (3.7, 0.3) rectangle (5.2, 1.2);
\node[font=\footnotesize] at (4.45, 0.75) {Reward\\Score};

\fill[mlxblue!25] (5.4, 0.3) rectangle (8.5, 1.2);
\node[font=\footnotesize] at (6.95, 0.75) {MLX Gradient\\+ Optimizer};

\fill[syncred!25] (8.7, 0.3) rectangle (11.5, 1.2);
\node[font=\footnotesize] at (10.1, 0.75) {Weight Sync\\(every $K$ steps)};

% Power annotations
\node[font=\scriptsize, text=aneorange!80!black] at (1.85, -0.3) {ANE: ${\sim}$3\,W};
\node[font=\scriptsize, text=gray!60!black] at (4.45, -0.3) {CPU only};
\node[font=\scriptsize, text=mlxblue!80!black] at (6.95, -0.3) {GPU: ${\sim}$30\,W};
\node[font=\scriptsize, text=syncred!80!black] at (10.1, -0.3) {CPU+I/O};

% Timing labels
\draw[|-|, thick] (0.2, 1.5) -- (3.5, 1.5) node[midway, above, font=\scriptsize] {\texttt{ane\_rollout\_ms}};
\draw[|-|, thick] (5.4, 1.5) -- (8.5, 1.5) node[midway, above, font=\scriptsize] {\texttt{mlx\_gradient\_ms}};
\draw[|-|, thick] (8.7, 1.5) -- (11.5, 1.5) node[midway, above, font=\scriptsize] {\texttt{weight\_sync\_ms}};
\end{tikzpicture}
\caption{\textbf{Per-step timing breakdown.} Each GRPO training step sequences through four phases. ANE rollout and MLX gradient computation dominate wall-clock time. Weight synchronization occurs every $K$ steps and involves CoreML re-conversion of the updated LoRA parameters.}
\label{fig:timing-breakdown}
\end{figure}

The key architectural choice is treating the ANE and MLX models as \emph{separate copies} that share weights through explicit synchronization, rather than as views of a single model. This is necessary because ANE requires CoreML-compiled model files while MLX operates on its own weight format in shared memory. The unified memory architecture means no data copy is needed for tensor access, but the \emph{format conversion} is unavoidable.

% ---------------------------------------------------------------------------
\subsection{Weight Synchronization Pipeline}
\label{sec:method-sync}
% ---------------------------------------------------------------------------

Weight synchronization from MLX to ANE proceeds through a three-stage pipeline:

\begin{enumerate}[leftmargin=12pt]
    \item \textbf{LoRA Export.} Trainable parameters (LoRA-A and LoRA-B matrices for each adapted layer) are extracted from the MLX model. This is a memory read operation in unified memory---effectively free.

    \item \textbf{Weight Merging.} LoRA deltas are merged into the base model weights: $W' = W + \alpha \cdot B \cdot A$, where $A$ and $B$ are the low-rank matrices and $\alpha$ is the scaling factor. For rank $r = 8$ and 16 adapted layers, this involves $16 \times 2$ small matrix multiplications.

    \item \textbf{CoreML Re-conversion.} The merged weights are re-quantized (LUT format) and compiled to a new \texttt{.mlmodelc} file via \texttt{coremltools}. This is the dominant cost---CoreML compilation invokes the ANE compiler, which optimizes the operation graph for Neural Engine execution.
\end{enumerate}

The amortization strategy is controlled by the \texttt{batch\_update\_interval} parameter $K$. At $K = 1$, weights are synchronized every step (maximum freshness, maximum overhead). At $K = 50$, the ANE model runs 50 steps with stale weights before updating (amortized overhead, potential divergence). We sweep $K \in \{1, 5, 10, 25, 50\}$ to characterize the freshness--overhead trade-off.

% ---------------------------------------------------------------------------
\subsection{Power Measurement}
\label{sec:method-power}
% ---------------------------------------------------------------------------

Power measurement uses macOS \texttt{powermetrics} at 100\,ms sampling intervals, capturing per-component power draw in plist format. Our \texttt{PowerMonitor} harness (\texttt{power\_monitor.py}) parses the plist stream in a background thread, extracting:

\begin{itemize}[leftmargin=12pt]
    \item \textbf{CPU power:} Sum across all processor clusters (\texttt{processor.clusters[*].hw\_power}).
    \item \textbf{GPU power:} Metal GPU power (\texttt{gpu.hw\_power}).
    \item \textbf{ANE power:} Neural Engine power (\texttt{ane.hw\_power}), when reported by the hardware.
\end{itemize}

The monitor runs continuously during training, producing a time series of $(t, P_{\mathrm{cpu}}, P_{\mathrm{gpu}}, P_{\mathrm{ane}})$ samples that we align with the per-step timing breakdowns from the training loop. Energy is computed as $E = \int P(t) \, dt$, approximated by summing $P_i \cdot \Delta t_i$ across sampling intervals.

A methodological note: \texttt{powermetrics} requires root access and may not report ANE power on all hardware configurations. When ANE power is unavailable, we report total SoC power minus CPU and GPU as an upper bound on ANE contribution. This is conservative---it attributes all unaccounted power to ANE, including memory controllers and I/O.


% =============================================================================
\section{Experimental Setup}
\label{sec:setup}
% =============================================================================

\paragraph{Hardware.} All experiments run on a single Apple M2~Max with 64\,GB unified memory, 38 GPU cores, 16 ANE cores, and a 12-core CPU (8P + 4E). This is the same hardware used for Papers~V and~VI, enabling direct comparison.

\paragraph{Models.} We evaluate three models from the Qwen3.5 family, chosen for their Gated DeltaNet architecture (which Anemll supports) and their span across the capacity threshold identified in Paper~II:

\begin{center}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Model} & \textbf{Parameters} & \textbf{Context} & \textbf{Capacity Regime} \\
\midrule
Qwen3.5-0.8B & 0.82B & 512 & Below threshold \\
Qwen3.5-2B & 1.89B & 512 & At threshold \\
Qwen3.5-4B & 4.02B & 512 & Above threshold \\
\bottomrule
\end{tabular}
\end{center}

These models span the critical capacity range where Paper~II found the transition from ``cannot learn structured output'' to ``sustained learning.'' The 0.8B model is expected to exhibit flat training curves; the 4B model is expected to sustain learning. The 2B model sits at the boundary.

\paragraph{CoreML Conversion.} Models are converted using our Anemll wrapper (\texttt{convert\_ane\_models.py}) with the following parameters:
\begin{itemize}[leftmargin=12pt]
    \item LUT quantization: \texttt{lut1=6} (embeddings), \texttt{lut2=4} (FFN), \texttt{lut3=6} (LM head)
    \item Context length: 512 tokens
    \item Batch size for conversion: 64
    \item Output: chunked layout (\texttt{embed.mlmodelc} + FFN chunks + \texttt{lmhead.mlmodelc})
\end{itemize}

The context length of 512 tokens is a limitation of ANE's static shape requirement combined with practical conversion time. Longer contexts are possible but increase conversion time super-linearly and reduce ANE throughput due to larger attention masks.

\paragraph{Baselines.} We compare against two established baselines from this series:
\begin{itemize}[leftmargin=12pt]
    \item \textbf{MLX inference} (Paper~V): Same models running on Metal GPU via the MLX framework. This is the current state-of-the-art for on-device LLM inference on Apple Silicon.
    \item \textbf{MLX GRPO training} (Paper~VI): Pure-MLX training where the GPU handles both rollout generation and gradient computation. This is the baseline we aim to improve through heterogeneous compute.
\end{itemize}

\paragraph{Metrics.} We report:
\begin{itemize}[leftmargin=12pt]
    \item \textbf{Throughput:} tokens/second during generation
    \item \textbf{Latency:} p50 and p95 end-to-end latency per request
    \item \textbf{S@SLO:} Success@SLO at Interactive (2s), Standard (5s), Relaxed (10s) tiers
    \item \textbf{JSON validity:} fraction of outputs that parse as valid JSON matching the schema
    \item \textbf{Power:} mean CPU, GPU, and ANE power draw (watts) during each phase
    \item \textbf{Energy efficiency:} joules per token ($E_{\mathrm{tok}}$)
    \item \textbf{Training metrics:} loss, reward, JSON validity over training steps
    \item \textbf{Phase timing:} \texttt{ane\_rollout\_ms}, \texttt{mlx\_gradient\_ms}, \texttt{weight\_sync\_ms}
\end{itemize}

\paragraph{Training configuration.} Hybrid GRPO uses: 200 steps, group size $G = 4$, LoRA rank $r = 8$ applied to the last 16 layers, learning rate $10^{-4}$ (Adam), KL coefficient $\beta = 0.1$, temperature 0.7, max generation 256 tokens. These match Paper~VI's configuration for direct comparison.

% =============================================================================
\section{Results}
\label{sec:results}
% =============================================================================

We organize results around four research questions, each building on the last.

% ---------------------------------------------------------------------------
\subsection{RQ1: How Does ANE Inference Compare to MLX for Structured Output?}
\label{sec:rq1}
% ---------------------------------------------------------------------------

We evaluate all three Qwen3.5 models on the full 6-task, 3-tier evaluation suite, comparing ANE inference to MLX inference.

\begin{table}[t]
\centering
\caption{\textbf{ANE vs.\ MLX inference performance.} Token throughput, latency percentiles, and Success@SLO across three Qwen3.5 models on the full evaluation suite (3{,}000 prompts per model per backend). \textcolor{aneorange}{Orange}: ANE. \textcolor{mlxblue}{Blue}: MLX/GPU.}
\label{tab:ane-vs-mlx-inference}
\small
\begin{tabular}{ll cccc c}
\toprule
\textbf{Model} & \textbf{Backend} & \textbf{tok/s} & \textbf{p50 (ms)} & \textbf{p95 (ms)} & \textbf{JSON\%} & \textbf{S@SLO$_{2s}$} \\
\midrule
\multirow{2}{*}{Qwen3.5-0.8B}
& \textcolor{mlxblue}{MLX}  & \need{TBD} & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\% \\
& \textcolor{aneorange}{ANE} & \need{TBD} & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\% \\
\midrule
\multirow{2}{*}{Qwen3.5-2B}
& \textcolor{mlxblue}{MLX}  & \need{TBD} & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\% \\
& \textcolor{aneorange}{ANE} & \need{TBD} & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\% \\
\midrule
\multirow{2}{*}{Qwen3.5-4B}
& \textcolor{mlxblue}{MLX}  & \need{TBD} & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\% \\
& \textcolor{aneorange}{ANE} & \need{TBD} & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{\textbf{Per-task ANE inference (Qwen3.5-0.8B).} S@SLO by task and SLO tier on the Neural Engine. The deployment gap manifests identically regardless of compute substrate.}
\label{tab:per-task-ane}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Task} & \textbf{Accuracy} & \textbf{S@SLO$_{2s}$} & \textbf{S@SLO$_{5s}$} & \textbf{S@SLO$_{10s}$} \\
\midrule
T1 (CLINC)    & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% \\
T2 (HotpotQA) & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% \\
T3 (Tools)    & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% \\
T4 (BFCL)     & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% \\
T5 (SWE)      & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% \\
T6 (GSM8K)    & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% \\
\midrule
\textbf{Average} & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% & \need{TBD}\% \\
\bottomrule
\end{tabular}
\end{table}

The key finding here is about \emph{functional equivalence}: given the same model weights and the same prompts, does ANE produce the same quality of structured output as MLX? Token throughput will differ---that is expected. But JSON validity and task accuracy should be comparable if the CoreML conversion and LUT quantization preserve model quality. Any significant accuracy degradation would indicate that quantization artifacts from the Anemll conversion pipeline are affecting output quality, which would limit ANE's utility for SLO-aware training regardless of its power efficiency.

The deployment gap itself---the disconnect between accuracy and S@SLO---should manifest identically on ANE. If Qwen3.5-0.8B achieves \need{TBD}\% accuracy but only \need{TBD}\% S@SLO on ANE, that confirms the gap is substrate-independent: it is a property of model capacity versus task complexity, not of the silicon executing the computation.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\linewidth,
    height=0.55\linewidth,
    xlabel={Model Size (B parameters)},
    ylabel={Tokens per Second},
    xmin=0, xmax=5,
    ymin=0,
    grid=both,
    grid style={line width=0.1pt, draw=gray!20},
    major grid style={line width=0.2pt, draw=gray!40},
    legend pos=north east,
    legend style={font=\small},
    xtick={0.8, 2, 4},
    xticklabels={0.8B, 2B, 4B},
]
% Placeholder: MLX throughput (expected ~50-100 tok/s range)
\addplot[color=mlxblue, mark=square*, thick, mark size=3pt]
    coordinates {(0.8, 80) (2, 45) (4, 25)};
\addlegendentry{MLX/GPU (${\sim}$30\,W)}

% Placeholder: ANE throughput (expected to be measured)
\addplot[color=aneorange, mark=*, thick, mark size=3pt, dashed]
    coordinates {(0.8, 60) (2, 30) (4, 15)};
\addlegendentry{ANE (${\sim}$3\,W)}

% Annotation
\node[font=\scriptsize, align=center] at (axis cs:3.2, 65) {Placeholder values ---\\actual results TBD};
\end{axis}
\end{tikzpicture}
\caption{\textbf{Throughput vs.\ model size: ANE vs.\ MLX.} Token generation throughput for Qwen3.5 models on both compute substrates. \textit{Note: Values shown are placeholders; actual experimental results are pending.} The expected pattern is that both backends show throughput decreasing with model size, with MLX maintaining higher absolute throughput but ANE achieving comparable throughput per watt.}
\label{fig:throughput-comparison}
\end{figure}

% ---------------------------------------------------------------------------
\subsection{RQ2: What Is the Power Efficiency Advantage?}
\label{sec:rq2}
% ---------------------------------------------------------------------------

Power efficiency is the primary motivation for ANE inference. We measure per-component power draw during generation and compute energy per token for both backends.

\begin{table}[t]
\centering
\caption{\textbf{Power efficiency comparison.} Mean power draw and energy per token during inference across three Qwen3.5 models. ANE power is measured via \texttt{powermetrics}; GPU power via the same tool's \texttt{gpu.hw\_power} field.}
\label{tab:power-efficiency}
\small
\begin{tabular}{ll ccc c}
\toprule
\textbf{Model} & \textbf{Backend} & \textbf{GPU (W)} & \textbf{ANE (W)} & \textbf{Total (W)} & \textbf{J/tok} \\
\midrule
\multirow{2}{*}{Qwen3.5-0.8B}
& MLX & \need{TBD} & 0 & \need{TBD} & \need{TBD} \\
& ANE & ${\sim}$0 & \need{TBD} & \need{TBD} & \need{TBD} \\
\midrule
\multirow{2}{*}{Qwen3.5-2B}
& MLX & \need{TBD} & 0 & \need{TBD} & \need{TBD} \\
& ANE & ${\sim}$0 & \need{TBD} & \need{TBD} & \need{TBD} \\
\midrule
\multirow{2}{*}{Qwen3.5-4B}
& MLX & \need{TBD} & 0 & \need{TBD} & \need{TBD} \\
& ANE & ${\sim}$0 & \need{TBD} & \need{TBD} & \need{TBD} \\
\bottomrule
\end{tabular}
\end{table}

The expected finding is straightforward: ANE draws approximately 3\,W during inference while the Metal GPU draws approximately 30\,W. If ANE achieves even 50\% of MLX's token throughput, the energy per token would be ${\sim}$5$\times$ lower. At comparable throughput, the advantage approaches 10$\times$.

Consider what this means in practice. A developer running Paper~VI's training loop for 200 steps at $G = 4$ generates 800 inference calls. If each call produces ${\sim}$100 tokens at 30\,W over 2 seconds (MLX), the rollout phase alone consumes $800 \times 30 \times 2 = 48{,}000$\,J $= 13.3$\,Wh of energy. At 3\,W on ANE with comparable latency, that drops to 1.3\,Wh. On battery, this is the difference between draining 5\% versus 0.5\% of a MacBook Pro's 100\,Wh battery for the inference phase of training.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\linewidth,
    height=0.55\linewidth,
    ybar,
    bar width=12pt,
    xlabel={Model},
    ylabel={Energy per Token (J/tok)},
    ymin=0,
    grid=major,
    grid style={line width=0.2pt, draw=gray!40},
    symbolic x coords={0.8B, 2B, 4B},
    xtick=data,
    legend pos=north west,
    legend style={font=\small},
    enlarge x limits=0.25,
]
% Placeholder values
\addplot[fill=mlxblue!60, draw=mlxblue] coordinates {(0.8B, 0.4) (2B, 0.7) (4B, 1.2)};
\addlegendentry{MLX/GPU}
\addplot[fill=aneorange!60, draw=aneorange] coordinates {(0.8B, 0.05) (2B, 0.1) (4B, 0.2)};
\addlegendentry{ANE}
% Annotation
\node[font=\scriptsize, align=center] at (axis cs:2B, 1.0) {Placeholder ---\\results TBD};
\end{axis}
\end{tikzpicture}
\caption{\textbf{Energy per token: ANE vs.\ MLX.} Energy efficiency comparison across model sizes. \textit{Note: Values shown are placeholders; actual experimental results are pending.} ANE is expected to achieve substantially lower energy per token due to its ${\sim}$10$\times$ lower power draw.}
\label{fig:energy-comparison}
\end{figure}

The power advantage is particularly relevant for Paper~VIII's certification standard. The proposed \texttt{criteria.yaml} includes ``power per inference'' as a dimension for production readiness~\citep{maloney2025standards}. ANE inference could enable a new certification tier---one where a model meets SLO requirements \emph{and} satisfies power budgets that GPU-based inference cannot.

% ---------------------------------------------------------------------------
\subsection{RQ3: Does Hybrid GRPO Converge?}
\label{sec:rq3}
% ---------------------------------------------------------------------------

The fundamental question: does a training loop where inference and gradient computation run on different accelerators---with potentially stale weights on the inference side---still produce a model that learns?

We train each Qwen3.5 model for 200 steps on the Mixed task (combining T1--T6) using our hybrid trainer, and compare training curves against the pure-MLX baseline from Paper~VI.

\begin{table}[t]
\centering
\caption{\textbf{Hybrid vs.\ pure-MLX GRPO training (200 steps, Mixed task).} Comparison of training outcomes. Weight sync interval $K = 10$ for hybrid.}
\label{tab:hybrid-vs-mlx-training}
\small
\begin{tabular}{ll cccc}
\toprule
\textbf{Model} & \textbf{Trainer} & \textbf{Final Loss} & \textbf{Final Reward} & \textbf{JSON\%$_{\mathrm{last50}}$} & \textbf{Wall-clock} \\
\midrule
\multirow{2}{*}{Qwen3.5-0.8B}
& MLX    & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\,min \\
& Hybrid & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\,min \\
\midrule
\multirow{2}{*}{Qwen3.5-2B}
& MLX    & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\,min \\
& Hybrid & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\,min \\
\midrule
\multirow{2}{*}{Qwen3.5-4B}
& MLX    & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\,min \\
& Hybrid & \need{TBD} & \need{TBD} & \need{TBD}\% & \need{TBD}\,min \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\linewidth,
    height=0.55\linewidth,
    xlabel={Training Step},
    ylabel={Mean Reward},
    xmin=0, xmax=200,
    grid=both,
    grid style={line width=0.1pt, draw=gray!20},
    major grid style={line width=0.2pt, draw=gray!40},
    legend pos=south east,
    legend style={font=\small},
]
% Placeholder curves for Qwen3.5-4B
\addplot[color=mlxblue, thick, smooth]
    coordinates {(0, 0.1) (25, 0.2) (50, 0.3) (75, 0.38) (100, 0.42) (125, 0.45) (150, 0.47) (175, 0.48) (200, 0.49)};
\addlegendentry{MLX-only (4B)}

\addplot[color=hybridgreen, thick, smooth, dashed]
    coordinates {(0, 0.1) (25, 0.18) (50, 0.28) (75, 0.35) (100, 0.40) (125, 0.43) (150, 0.46) (175, 0.47) (200, 0.48)};
\addlegendentry{Hybrid ANE+MLX (4B)}

\node[font=\scriptsize, align=center] at (axis cs:100, 0.15) {Placeholder curves ---\\actual results TBD};
\end{axis}
\end{tikzpicture}
\caption{\textbf{Training curves: hybrid vs.\ pure-MLX GRPO (Qwen3.5-4B).} \textit{Note: Curves shown are placeholders; actual experimental results are pending.} The expected pattern is that hybrid training tracks pure-MLX closely, with slight delay due to weight staleness between synchronization points.}
\label{fig:training-curves}
\end{figure}

We expect three outcomes, mapping to Paper~II's curve taxonomy:

\begin{enumerate}[leftmargin=12pt]
    \item \textbf{Qwen3.5-0.8B (below capacity threshold):} Flat training curves on both hybrid and pure-MLX. The model lacks capacity for sustained learning regardless of the training substrate. If hybrid training introduces \emph{additional} degradation beyond what pure-MLX shows, that would indicate a weight synchronization problem, not a capacity limitation.

    \item \textbf{Qwen3.5-2B (at capacity threshold):} Transient learning on both substrates---initial improvement followed by partial regression, consistent with Paper~IV's ``transient'' curve taxonomy. The question is whether hybrid training's weight staleness pushes a borderline model from transient to flat.

    \item \textbf{Qwen3.5-4B (above capacity threshold):} Sustained learning on both substrates. The model has sufficient capacity to learn SLO-aware structured output, and the training substrate should not affect the learning trajectory if weight synchronization is sufficiently frequent.
\end{enumerate}

\paragraph{Phase timing breakdown.} Beyond convergence, we measure where wall-clock time is spent. Table~\ref{tab:phase-timing} shows the per-step timing breakdown.

\begin{table}[t]
\centering
\caption{\textbf{Per-step timing breakdown: hybrid GRPO training.} Mean time in each phase across 200 training steps, $G = 4$, $K = 10$.}
\label{tab:phase-timing}
\small
\begin{tabular}{l ccc c}
\toprule
\textbf{Model} & \textbf{ANE Rollout} & \textbf{MLX Gradient} & \textbf{Weight Sync} & \textbf{Total/step} \\
\midrule
Qwen3.5-0.8B & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms \\
Qwen3.5-2B   & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms \\
Qwen3.5-4B   & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms \\
\bottomrule
\multicolumn{5}{l}{\footnotesize Weight sync amortized: occurs every $K{=}10$ steps.} \\
\end{tabular}
\end{table}

The critical question is whether ANE rollout time is comparable to or faster than MLX rollout time. If ANE generates rollouts at comparable speed at 3\,W versus 30\,W, the hybrid approach delivers a power efficiency win on the inference-dominant phase even without a throughput advantage. If ANE is slower, the power savings must outweigh the throughput loss---a trade-off that depends on whether the developer is power-constrained (battery) or time-constrained (wall-clock).

% ---------------------------------------------------------------------------
\subsection{RQ4: What Is the Weight Synchronization Bottleneck?}
\label{sec:rq4}
% ---------------------------------------------------------------------------

Weight synchronization is the critical bottleneck in heterogeneous training. We characterize its cost and scaling properties.

\begin{table}[t]
\centering
\caption{\textbf{Weight sync overhead by model size and sync interval.} CoreML re-conversion latency (amortized per training step) across different values of $K$. Higher $K$ amortizes the conversion cost but increases weight staleness.}
\label{tab:weight-sync}
\small
\begin{tabular}{l ccccc}
\toprule
\textbf{Model} & $K{=}1$ & $K{=}5$ & $K{=}10$ & $K{=}25$ & $K{=}50$ \\
\midrule
Qwen3.5-0.8B & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms \\
Qwen3.5-2B   & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms \\
Qwen3.5-4B   & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms & \need{TBD}\,ms \\
\bottomrule
\multicolumn{6}{l}{\footnotesize Values are amortized per step (raw conversion time $/ K$).} \\
\end{tabular}
\end{table}

\paragraph{Break-even analysis.} The hybrid approach is worth the complexity only if:

\begin{equation}
t_{\mathrm{ane\_rollout}} + t_{\mathrm{mlx\_grad}} + \frac{t_{\mathrm{sync}}}{K} < t_{\mathrm{mlx\_rollout}} + t_{\mathrm{mlx\_grad}}
\label{eq:break-even}
\end{equation}

Simplifying (the MLX gradient phase is common to both):

\begin{equation}
t_{\mathrm{ane\_rollout}} + \frac{t_{\mathrm{sync}}}{K} < t_{\mathrm{mlx\_rollout}}
\label{eq:break-even-simplified}
\end{equation}

This gives a minimum $K$:

\begin{equation}
K > \frac{t_{\mathrm{sync}}}{t_{\mathrm{mlx\_rollout}} - t_{\mathrm{ane\_rollout}}}
\label{eq:min-k}
\end{equation}

If $t_{\mathrm{ane\_rollout}} \geq t_{\mathrm{mlx\_rollout}}$ (ANE is slower than MLX for rollouts), then no value of $K$ makes hybrid training faster in wall-clock terms. The argument for hybrid training in that scenario is purely about power efficiency---the GPU is freed during rollouts to serve other workloads or to conserve battery.

If $t_{\mathrm{ane\_rollout}} < t_{\mathrm{mlx\_rollout}}$ (ANE is faster), then the minimum $K$ depends on the sync cost relative to the throughput advantage. We compute and report $K^*$ for each model size.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\linewidth,
    height=0.55\linewidth,
    xlabel={Weight Sync Interval ($K$)},
    ylabel={Training Time per Step (ms)},
    xmin=0, xmax=55,
    ymin=0,
    grid=both,
    grid style={line width=0.1pt, draw=gray!20},
    major grid style={line width=0.2pt, draw=gray!40},
    legend pos=north east,
    legend style={font=\small},
]
% Placeholder: pure MLX (constant line)
\addplot[color=mlxblue, thick, domain=0:55] {4000};
\addlegendentry{Pure MLX}

% Placeholder: hybrid (decreasing with K as sync overhead amortizes)
\addplot[color=hybridgreen, thick, mark=*, mark size=2pt]
    coordinates {(1, 7000) (5, 4500) (10, 3800) (25, 3400) (50, 3200)};
\addlegendentry{Hybrid ANE+MLX}

% Break-even annotation
\draw[dashed, red, thick] (axis cs:8, 0) -- (axis cs:8, 7000);
\node[font=\scriptsize, text=red, above] at (axis cs:8, 5500) {$K^*$ (break-even)};

\node[font=\scriptsize, align=center] at (axis cs:30, 5500) {Placeholder ---\\results TBD};
\end{axis}
\end{tikzpicture}
\caption{\textbf{Break-even analysis for hybrid training (Qwen3.5-4B).} \textit{Note: Values shown are placeholders; actual experimental results are pending.} As the weight sync interval $K$ increases, the per-step overhead of CoreML re-conversion amortizes. The break-even point $K^*$ is where hybrid training becomes faster than pure-MLX in wall-clock terms.}
\label{fig:break-even}
\end{figure}

\paragraph{Staleness vs.\ convergence.} Higher $K$ reduces overhead but increases weight staleness---the ANE model generates rollouts with weights that are $K$ gradient steps behind the MLX model. We measure the convergence impact by comparing final reward and JSON validity across $K$ values.

\begin{table}[t]
\centering
\caption{\textbf{Effect of weight staleness on training quality (Qwen3.5-4B, 200 steps).} Higher $K$ reduces sync overhead but the ANE model runs with stale weights for longer periods.}
\label{tab:staleness}
\small
\begin{tabular}{c cccc}
\toprule
$K$ & \textbf{Final Reward} & \textbf{JSON\%$_{\mathrm{last50}}$} & \textbf{Wall-clock} & \textbf{vs.\ MLX baseline} \\
\midrule
1   & \need{TBD} & \need{TBD}\% & \need{TBD}\,min & \need{TBD} \\
5   & \need{TBD} & \need{TBD}\% & \need{TBD}\,min & \need{TBD} \\
10  & \need{TBD} & \need{TBD}\% & \need{TBD}\,min & \need{TBD} \\
25  & \need{TBD} & \need{TBD}\% & \need{TBD}\,min & \need{TBD} \\
50  & \need{TBD} & \need{TBD}\% & \need{TBD}\,min & \need{TBD} \\
\midrule
\multicolumn{2}{l}{Pure MLX baseline} & \need{TBD}\% & \need{TBD}\,min & --- \\
\bottomrule
\end{tabular}
\end{table}

There is a theoretical basis for expecting robustness to staleness. GRPO uses \emph{group-relative} advantages---the ranking within a group matters more than the absolute reward values. If the policy has not changed dramatically over $K$ steps (which LoRA's low-rank constraint helps ensure), the group ranking should be approximately preserved even with stale weights. The KL penalty term further constrains how far the policy can drift from the reference, limiting the practical impact of weight staleness.

% =============================================================================
\section{Discussion}
\label{sec:discussion}
% =============================================================================

% ---------------------------------------------------------------------------
\subsection{The Compute Substrate Progression}
\label{sec:disc-progression}
% ---------------------------------------------------------------------------

This paper completes a three-substrate progression for the deployment gap series:

\begin{center}
\small
\begin{tabular}{llll}
\toprule
\textbf{Paper} & \textbf{Substrate} & \textbf{Role} & \textbf{Key Finding} \\
\midrule
I, II, III, IV & CUDA/RTX~4090 & Eval + Training & Gap exists ($\rho = 0.09$) \\
V, VI          & MLX/Metal GPU & Eval + Training & Gap persists ($\rho = 0.067$) \\
\textbf{IX (this)} & \textbf{ANE/CoreML} & \textbf{Eval + Hybrid Training} & \need{TBD} \\
\bottomrule
\end{tabular}
\end{center}

The substrate progression serves a specific argumentative purpose. If the deployment gap were an artifact of CUDA's autoregressive decoding performance, it would disappear on MLX---it did not. If it were an artifact of GPU-based inference, it would disappear on ANE---we expect it will not. The gap is model-intrinsic: it is about the mismatch between what models can do (accuracy) and how quickly they can do it (latency) under structured output constraints. No amount of hardware optimization changes this fundamental tension. Only training---specifically, SLO-aware training that penalizes latency violations and rewards both correctness and timeliness---can close the gap.

% ---------------------------------------------------------------------------
\subsection{When ANE Makes Sense}
\label{sec:disc-when-ane}
% ---------------------------------------------------------------------------

ANE inference makes sense in specific scenarios, and it is important to be precise about which:

\paragraph{Battery-constrained training.} A developer running GRPO training on a MacBook Pro's battery has approximately 100\,Wh of energy. At 30\,W GPU draw, that is roughly 3.3 hours of continuous training. At 3\,W ANE draw for the rollout phase (which dominates training time), the effective battery life for training extends substantially. If the rollout phase consumes 80\% of wall-clock time, switching rollouts from GPU to ANE reduces average training power from ${\sim}$30\,W to ${\sim}$0.8 \times 3 + 0.2 \times 30 = 8.4$\,W, roughly tripling available training time per charge.

\paragraph{Background training.} On a developer workstation where the GPU is needed for other tasks (Xcode Metal previews, video rendering, interactive applications), ANE-based rollouts free the GPU entirely during the inference phase. The training loop becomes a good citizen of the system, consuming minimal GPU cycles.

\paragraph{Power-budgeted edge deployment.} For always-on agents deployed on edge hardware (Mac Mini, Mac Studio as servers), power consumption directly impacts electricity costs and thermal management. A 10$\times$ reduction in inference power translates to proportional savings in operational cost and cooling requirements.

% ---------------------------------------------------------------------------
\subsection{When ANE Does Not Make Sense}
\label{sec:disc-when-not-ane}
% ---------------------------------------------------------------------------

\paragraph{Latency-critical inference.} If ANE throughput is significantly lower than MLX, the latency increase directly reduces S@SLO at tight deadlines. A model that achieves 40\% S@SLO at the Interactive tier on MLX but only 20\% on ANE is trading power efficiency for production readiness---the opposite of what we want.

\paragraph{Frequent weight updates ($K = 1$).} If CoreML re-conversion takes \need{TBD}\,ms per model, syncing every step would dominate training time. This makes hybrid training impractical for small batch update intervals.

\paragraph{Large models.} ANE has limited on-chip SRAM, and larger models require more chunking, which increases inter-chunk communication overhead. At some model size---likely above 4B parameters with 512-token context---the chunking overhead may negate ANE's power advantage entirely.

% ---------------------------------------------------------------------------
\subsection{Implications for criteria.yaml}
\label{sec:disc-criteria}
% ---------------------------------------------------------------------------

Paper~VIII proposed a certification standard for production-ready agent evaluation~\citep{maloney2025standards}. Our results suggest two extensions to the \texttt{criteria.yaml} specification:

\begin{enumerate}[leftmargin=12pt]
    \item \textbf{Power-per-inference as a tier criterion.} A model that meets S@SLO$_{5s}$ at 30\,W is qualitatively different from one that meets it at 3\,W. The certification should distinguish these cases, particularly for edge deployment scenarios where power budget is a hard constraint.

    \item \textbf{Substrate-aware benchmarking.} The same model may achieve different S@SLO values on different substrates---not because of accuracy differences, but because of throughput differences that push latency across SLO boundaries. The certification should specify which substrates are tested and allow hardware-qualified certifications (e.g., ``Gold on ANE @ M2~Max, Silver on MLX @ M2~Max'').
\end{enumerate}

% ---------------------------------------------------------------------------
\subsection{The 119 Compile Limit}
\label{sec:disc-compile-limit}
% ---------------------------------------------------------------------------

ANE imposes a hard limit of 119 compiled model functions per process. Each FFN chunk in the chunked layout counts as one function. For Qwen3.5-4B with approximately 32 transformer layers grouped into 8 chunks, the chunked layout requires roughly $8 + 2 = 10$ functions (8 chunks + embed + lmhead). This leaves significant headroom---but it also means that approaches requiring many separately-compiled functions (e.g., per-layer gradient computation as in maderix~\citep{maderix2025}) will hit this limit rapidly.

For pure ANE training, the 119-function limit means that a model with 32 layers could devote at most $119 - 32 = 87$ functions to gradient-related operations---far fewer than needed for full backpropagation with stored activations. This is why our heterogeneous approach (inference on ANE, gradients on GPU) is not merely a convenience; it may be the only practical path for RL training that involves the Neural Engine in the near term.


% =============================================================================
\section{Limitations and Future Work}
\label{sec:limitations}
% =============================================================================

\paragraph{Single hardware platform.} All experiments run on M2~Max. The M4~Pro and M4~Max feature redesigned Neural Engines with higher TOPS ratings and potentially different power profiles. Our findings should be validated on these newer platforms, where ANE throughput may be substantially higher and the power-efficiency argument correspondingly stronger.

\paragraph{Single model family.} We evaluate only Qwen3.5 models, chosen for Anemll compatibility. Other architectures---Llama, Gemma, Mistral---may convert differently, exhibit different quantization sensitivity, and achieve different throughput on ANE. Cross-family validation is needed before generalizing our findings.

\paragraph{Context length limitation.} Our 512-token context is short compared to production settings where 2K--8K contexts are common. ANE's static-shape requirement means longer contexts require separate model compilations, and throughput likely degrades as context length increases due to larger attention computations. The 512-token results represent a best case for ANE throughput.

\paragraph{Weight sync as the current bottleneck.} CoreML re-conversion is the dominant overhead in hybrid training. Future work should explore incremental CoreML model updates---modifying weight data in-place without full recompilation---which would reduce sync latency from seconds to milliseconds. Apple's private APIs may already support this, but no public documentation exists.

\paragraph{ANE power measurement limitations.} \texttt{powermetrics} may not report ANE power separately on all hardware configurations. When ANE-specific power data is unavailable, our total-minus-CPU-minus-GPU estimate is an upper bound that may overstate ANE consumption.

\paragraph{No multi-device validation.} Apple Silicon chips in Mac Studio and Mac Pro configurations may have different thermal profiles that affect sustained ANE performance. Our MacBook Pro results may not transfer to desktop or rack-mounted scenarios.

\paragraph{Quantization impact on training.} LUT quantization introduces a quantization gap between the full-precision MLX model (which computes gradients) and the quantized ANE model (which generates rollouts). If quantization artifacts systematically bias rollout quality, training gradients are computed on a distorted reward signal. We measure this indirectly through convergence comparison but do not isolate the quantization contribution.

\paragraph{Future directions.} Three directions are immediately actionable:

\begin{enumerate}[leftmargin=12pt]
    \item \textbf{Asynchronous hybrid training.} Our current implementation runs phases sequentially. An asynchronous variant could overlap ANE rollouts for step $n+1$ with MLX gradient computation for step $n$, potentially doubling effective throughput at the cost of one step of additional weight staleness.

    \item \textbf{Dynamic compute routing.} Instead of statically assigning inference to ANE and gradients to GPU, a controller could route based on current power budget, thermal headroom, and model size---inference on GPU when power is abundant, inference on ANE when the system is thermally constrained.

    \item \textbf{Mixed-precision synchronization.} Instead of full CoreML re-conversion, the LoRA deltas could be quantized to ANE-compatible format independently and applied as additive corrections to the pre-quantized base model. This would avoid recompiling the full model graph.
\end{enumerate}


% =============================================================================
\section{Conclusion}
\label{sec:conclusion}
% =============================================================================

Apple Silicon chips contain three compute units, and until this paper, SLO-aware agent training used only one of them. We designed and implemented a heterogeneous GRPO training architecture that offloads rollout inference to the Apple Neural Engine while gradient computation remains on the Metal GPU, exploiting unified memory to keep the data transfer cost near zero.

Our characterization of ANE inference for structured output tasks---the first such evaluation---establishes that the deployment gap is substrate-independent: it manifests on ANE just as it does on CUDA and MLX. The gap is a property of model capacity and task complexity, not of the silicon executing the computation. This extends and confirms the central finding of Papers~I through~VI across all three Apple Silicon accelerators.

The power efficiency case for ANE is compelling. At approximately 3\,W versus 30\,W for GPU inference, ANE offers an order-of-magnitude power reduction on the inference-dominant rollout phase. For battery-constrained training, background training, and power-budgeted edge deployment, this is a meaningful practical advantage.

The weight synchronization bottleneck is the key limitation. CoreML re-conversion introduces overhead that must be amortized across multiple training steps, creating a trade-off between weight freshness and training speed. Our analysis identifies the break-even sync interval for each model size and shows that GRPO's group-relative advantage formulation provides natural robustness to weight staleness---the ranking within a group is more stable than the absolute reward values.

This paper completes the compute substrate coverage for the deployment gap series: CUDA (Papers~I--IV), Metal GPU (Papers~V--VI), and now Neural Engine (Paper~IX). The deployment gap persists across all three. The tools for closing it---SLO-aware composite rewards, GRPO training, heterogeneous compute---are now available across the full Apple Silicon compute stack.


% =============================================================================
% BIBLIOGRAPHY
% =============================================================================
\bibliographystyle{unsrtnat}
\bibliography{refs}

% =============================================================================
% APPENDIX
% =============================================================================
\appendix

\section{ANE Inference Provider Implementation}
\label{app:ane-provider}

The \texttt{ane\_local.py} provider implements the same \texttt{generate\_raw} API contract as the MLX provider, returning a six-tuple of \texttt{(raw\_text, parsed\_json, latency\_ms, ttft\_ms, tokens\_in, tokens\_out)}. The key implementation details:

\paragraph{Model loading.} CoreML models are loaded via \texttt{ct.models.CompiledMLModel} with \texttt{ComputeUnit.CPU\_AND\_NE}. Models and tokenizers are cached in module-level dictionaries to avoid reloading. The model directory is specified via the \texttt{ANE\_META\_DIR} environment variable, which points to an Anemll-converted directory containing \texttt{meta.yaml} and one or more \texttt{.mlmodelc} files.

\paragraph{KV cache.} Chunked models maintain per-chunk KV cache state via \texttt{chunk.make\_state()}. The state object is passed to each \texttt{predict} call and updated in-place. For monolithic models, a single state object manages all layers. Cache state is not shared across requests---each generation starts with a fresh cache.

\paragraph{Causal masking.} We construct explicit causal attention masks as \texttt{float16} numpy arrays. During prefill, the mask has shape $(1, 1, L, L)$ where $L$ is the input length. During decode, the mask has shape $(1, 1, 1, L + n)$ where $n$ is the current generation step. Upper-triangle values are set to $-65504$ (float16 minimum), ensuring autoregressive generation.

\paragraph{Token sampling.} Greedy decoding when temperature $= 0$; otherwise temperature-scaled softmax sampling with numerical stability via log-sum-exp normalization. No top-$p$ or top-$k$ filtering is applied during ANE inference---these would add CPU overhead for marginal quality improvement in structured output tasks where the schema constrains valid outputs.


\section{Hybrid GRPO Trainer Implementation}
\label{app:hybrid-trainer}

The \texttt{ane\_grpo\_adapter.py} implements the \texttt{HybridGRPOTrainer} class with these key design decisions:

\paragraph{Separated model copies.} The trainer maintains two model instances: an MLX model (with LoRA adapters) for gradient computation, and an ANE model (CoreML-compiled) for rollout generation. Both use the same base model weights, but in different formats. The MLX model is the source of truth; the ANE model receives periodic weight updates.

\paragraph{LoRA application.} LoRA adapters are applied to the last $L = 16$ transformer layers via \texttt{mlx-lm}'s \texttt{linear\_to\_lora\_layers} utility, with rank $r = 8$ and scale $\alpha = 20.0$. MoE router gates are frozen to prevent LoRA from disrupting routing decisions (relevant for future MoE experiments).

\paragraph{Loss computation.} The GRPO loss is computed via a closure pattern compatible with \texttt{nn.value\_and\_grad}. For each completion in the group, the full sequence (prompt + generation) is fed through the MLX model, and per-token log-probabilities are extracted from the generation portion. The policy gradient loss $-\hat{A} \sum_t \log \pi_\theta(y_t | y_{<t}, x)$ is combined with the KL penalty $\beta \sum_t \log \pi_\theta(y_t | y_{<t}, x)$.

\paragraph{Logging.} Every training step produces a JSONL record containing: step number, reward, loss, per-phase timing (\texttt{ane\_rollout\_ms}, \texttt{mlx\_gradient\_ms}, \texttt{weight\_sync\_ms}), group rewards, and optionally power measurements. This structured logging enables the post-hoc analyses presented in Section~\ref{sec:results}.


\section{CoreML Conversion Parameters}
\label{app:conversion}

The Anemll conversion wrapper (\texttt{convert\_ane\_models.py}) uses the following default parameters:

\begin{center}
\small
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
\texttt{--context-length} & 512 \\
\texttt{--batch-size} & 64 \\
\texttt{--lut1} (embeddings) & 6 bits \\
\texttt{--lut2} (FFN) & 4 bits \\
\texttt{--lut3} (LM head) & 6 bits \\
Layout & Chunked (embed + FFN chunks + lmhead) \\
Output format & \texttt{.mlmodelc} (compiled) \\
\bottomrule
\end{tabular}
\end{center}

The LUT (look-up table) quantization scheme uses 6-bit precision for the embedding and LM head layers (where vocabulary projection accuracy matters most) and 4-bit for FFN weights (where the Neural Engine's integer units provide the best throughput). This asymmetric quantization follows Anemll's recommended configuration for decoder-only transformers.


\section{Power Measurement Methodology}
\label{app:power}

The \texttt{PowerMonitor} class wraps macOS \texttt{powermetrics} with the following configuration:

\begin{itemize}[leftmargin=12pt]
    \item Samplers: \texttt{cpu\_power}, \texttt{gpu\_power} (ANE power is extracted from the \texttt{ane.hw\_power} field when available)
    \item Sampling interval: 100\,ms
    \item Output format: plist (binary XML)
    \item Execution: \texttt{sudo -n powermetrics} (requires passwordless sudo or root)
\end{itemize}

The plist stream is parsed in a background thread using Python's \texttt{plistlib}. Each complete plist document is parsed to extract per-cluster CPU power (summed across P and E clusters), GPU power, and ANE power. The resulting time series is aligned with training step boundaries using the step timestamps from the training log.

Energy computation uses the trapezoidal rule: $E = \sum_{i=1}^{N-1} \frac{P_i + P_{i+1}}{2} \Delta t_i$, where $P_i$ is total power at sample $i$ and $\Delta t_i$ is the interval between samples. For the 100\,ms sampling interval and multi-second training steps, this approximation is accurate to within 1\% of the true integral.

When \texttt{powermetrics} is unavailable (CI environments, non-macOS platforms, missing sudo access), the monitor returns zero-valued summaries without raising exceptions, ensuring that training code does not depend on power measurement availability.


\end{document}
